```{r}
# ============================================================
# Title:    Position P1 Statistical Analysis Script
# Project:  Individual Differences of Inhibition Control in Bilingual Sentence Processing: A Self Paced Reading Approach
# OSF:      https://osf.io/uyvxw
# Author:   Niranjana Hegde Bhimanakone Satyanarayana
# Created:  28/10/2025
# R version: 4.3.2
#
# Description:
# This script uses linear-mixed effects models to analyse the self-paced reading times
# at position P1. The statistical models at this position investigates our RQ1a i.e.,
# "How is inhibition control of L1 representations manifested in L2 sentence reading,
# as reflected in self-paced reading times?"
# The predictor variable 'inhibition_control_score' tests the RQ2,
# "Do the individual differences of interference and inhibition control modulate
# the reading times of interlingual homographs and their translation equivalent words within a sentence?"
#
# Note:
# In this file, following shorthand notations are used:
#       - Base - Confirmatory Model
#       - Exp  - Exploratory Model
#       - ROI  - Region of interest where the target words are placed
#       - SO1  - Spillover 1
#       - SO2  - Spillover 2
# ============================================================
```
```{r}
# ============================================================
# Roadmap:
#   - The confirmatory model analyses test only the effects that are
#     specified a priori and preregistered on OSF, namely the effects
#     of the P1 and P2 word types and inhibition control, while also including standard
#     control covariates such as word length and log-transformed word frequency.
#   - The exploratory model analyses extend the confirmatory models by adding further covariates such as
#     orthographic distance, phonetic distance, and LexTALE scores to examine whether they
#     had effects of their own and whether they explained variance previously captured by the
#     preregistered predictors
#   - At each step, predictors are removed individually,
#     and the resulting reduced models were compared to the full model using
#     ANOVA. Additionally, in the case of singularity, the random effects are simplified
#   - Above steps are applicable to all regions and to both confirmatory and exploratory models
# ============================================================
```
```{r}
# Importing the libraries
library(ggeffects)
library(tidyr)
library(lmerTest)
library(ggplot2)
library(dplyr)
conflicted::conflicts_prefer(dplyr::select)
conflicted::conflicts_prefer(dplyr::rename)
conflicted::conflicts_prefer(dplyr::mutate)
conflicted::conflicts_prefer(dplyr::filter)
```
```{r}
# Results files and SUBTLEX files
spr_file_list1 <- "D:/Studies/Thesis/Analysis/Inhibition-Control-Bilingual-Sentence-Processing/data/spr/29107b08-8dc4-4987-b1a3-1f0596ac3c7e_results.csv"
spr_file_list2 <- "D:/Studies/Thesis/Analysis/Inhibition-Control-Bilingual-Sentence-Processing/data/spr/06a109e9-1300-45b0-8445-c45643f19279_results.csv"
spr_file_list3 <- "D:/Studies/Thesis/Analysis/Inhibition-Control-Bilingual-Sentence-Processing/data/spr/25e35492-fc79-4a22-bcc5-7f7d89a9aee8_results.csv"
spr_file_list4 <- "D:/Studies/Thesis/Analysis/Inhibition-Control-Bilingual-Sentence-Processing/data/spr/eed95789-5869-4db7-8a1c-210b900bcd14_results.csv"
final_participants_file <- "D:/Studies/Thesis/Analysis/Inhibition-Control-Bilingual-Sentence-Processing/data/final analysis/inhibition_control_composite_scores.csv"
list1 <- "D:/Studies/Thesis/Analysis/Inhibition-Control-Bilingual-Sentence-Processing/materials/List1.csv"
list2 <- "D:/Studies/Thesis/Analysis/Inhibition-Control-Bilingual-Sentence-Processing/materials/List2.csv"
list3 <- "D:/Studies/Thesis/Analysis/Inhibition-Control-Bilingual-Sentence-Processing/materials/List3.csv"
list4 <- "D:/Studies/Thesis/Analysis/Inhibition-Control-Bilingual-Sentence-Processing/materials/List4.csv"
materials_file <- "D:/Studies/Thesis/Analysis/Inhibition-Control-Bilingual-Sentence-Processing/materials/FinalMaterials.csv"
inhibition_composite_score_file <- "D:/Studies/Thesis/Analysis/Inhibition-Control-Bilingual-Sentence-Processing/data/final analysis/inhibition_control_composite_scores.csv"
item_properties_file <- "D:/Studies/Thesis/Analysis/Inhibition-Control-Bilingual-Sentence-Processing/materials/itemsWithDistances.csv"
lexTALE_file <- "D:/Studies/Thesis/Analysis/Inhibition-Control-Bilingual-Sentence-Processing/data/final analysis/participants_based_on_proficiency.csv"
subtlex_us_file <- "D:\\Studies\\Thesis\\Analysis\\Inhibition-Control-Bilingual-Sentence-Processing\\materials\\SUBTLEXus.Rdata"
load(subtlex_us_file)
```
```{r}
## Read all Materials
spr_df_l1 <- read.csv(spr_file_list1)
spr_df_l2 <- read.csv(spr_file_list2)
spr_df_l3 <- read.csv(spr_file_list3)
spr_df_l4 <- read.csv(spr_file_list4)
final_participants_df <- read.csv(final_participants_file)
l1_materials <- read.csv(list1)
l2_materials <- read.csv(list2)
l3_materials <- read.csv(list3)
l4_materials <- read.csv(list4)
materials_df <- read.csv(materials_file)
item_properties_df <- read.csv(item_properties_file)
lexTale_df <- read.csv(lexTALE_file)
inhibition_composite_score_df <- read.csv(inhibition_composite_score_file)
```
```{r}
# Bind the response files together
spr_df_l3$Newline. <- as.character(spr_df_l3$Newline.) # This line is added because I mistakenly changed the column datatype of this column when I opened it in Excel.
spr_df <- bind_rows(spr_df_l1, spr_df_l2, spr_df_l3, spr_df_l4)

# Combine all materials, rename the column, and remove out fillers
materials_df <- bind_rows(l1_materials, l2_materials, l3_materials, l4_materials)
materials_df <- materials_df %>%
    filter(Type != "Filler") %>%
    rename(
        WordatP1 = P1
    )
```
```{r}
# Filter only those who are eligible from proficiency measures and inhibition control scores
spr_df <- spr_df %>%
    filter(user_id %in% inhibition_composite_score_df$workerid) %>%
    filter(user_id %in% lexTale_df$workerid[lexTale_df$comprehensionAccuracy > 0.50])
```
```{r}
# Filter only DashedSentences
spr_df_filtered <- spr_df %>%
    filter(PennElementName == "DashedSentence") %>%
    filter(Condition != "Filler" & Condition != "") # Remove fillers

spr_df_filtered <- spr_df_filtered %>%
    mutate(
        WordTypeP1 = as.factor(ifelse(grepl("IH", Condition), "IH", "C1")),
        WordTypeP2 = as.factor(ifelse(grepl("TE", Condition), "TE", "C2")),
    )
spr_df_filtered$ItemID <- as.integer(spr_df_filtered$ItemID)
```
```{r}
# Treat spr_df_filtered as base dataframe that can be reused even for spillover analyses
# So add all other info into the same dataframe for reusing
# Item Properties dataframe has just 24 rows, and all the C1 items are associated with IH for easier association.
# But now, I need them separately and using join will not help me extract different frequency values for IH and C1.
# So using transmute, and bind_rows to actually create a new lookup df altogether and later I can join them based on ItemID and P1.
lookup_df <- item_properties_df %>%
    transmute(
        ItemID,
        WordTypeP1 = "IH",
        LogFreqUS = LogFreqUS,
        LogFreqDE = LogFreqDE,
        phonetic_distance = Phonetic.Distance.Normalised,
        orthographic_distance = Orthographic.Distance.Normalised
    ) %>%
    bind_rows(
        item_properties_df %>%
            transmute(
                ItemID,
                WordTypeP1 = "C1",
                LogFreqUS = LogFreqC1,
                LogFreqDE = LogFreqDE,
                phonetic_distance = Phonetic.Distance.Normalised,
                orthographic_distance = Orthographic.Distance.Normalised
            )
    )

spr_df_filtered <- spr_df_filtered %>%
    left_join(lookup_df, by = c("ItemID", "WordTypeP1"))
```
```{r}
# Include individual inhibition control composite scores
inhibition_composite_score_df <- inhibition_composite_score_df %>% rename(user_id = workerid)
spr_df_filtered <- spr_df_filtered %>%
    left_join(
        inhibition_composite_score_df %>%
            select(user_id, composite_score, squared_simon_accuracy, squared_flanker_accuracy, squared_stroop_accuracy),
        by = "user_id"
    ) %>%
    rename(inhibition_control_score = composite_score)
```
```{r}
lexTale_df <- lexTale_df %>% rename(user_id = workerid)
spr_df_filtered <- spr_df_filtered %>%
    left_join(lexTale_df %>% select(user_id, score), by = "user_id") %>%
    rename(LexTALE = score)
```
``````{r}
# Remove Item 18 due to copy-paste error leading to multiple lists having the same condition
spr_df_filtered <- spr_df_filtered[spr_df_filtered$ItemID != 18, ]
```
```{r}
# Summary
summary(spr_df_filtered$squared_stroop_accuracy)
sd(spr_df_filtered$squared_stroop_accuracy)


summary(spr_df_filtered$squared_simon_accuracy)
sd(spr_df_filtered$squared_simon_accuracy)

summary(spr_df_filtered$squared_flanker_accuracy)
sd(spr_df_filtered$squared_flanker_accuracy)
```
```{r}
# Factorize participant and item variable
spr_df_filtered$user_id <- as.factor(spr_df_filtered$user_id)
spr_df_filtered$ItemID <- as.factor(spr_df_filtered$ItemID)
```
```{r}
# ============================================================
# 1. ROI Analysis
# ============================================================
```
```{r}
spr_df_for_P1 <- spr_df_filtered %>%
    left_join(materials_df %>% select(SentenceID, Critical.Position.Index, WordatP1),
        by = "SentenceID"
    ) %>%
    filter(Parameter == Critical.Position.Index)
```
```{r}
# Remove responses lesser than 100ms
original <- nrow(spr_df_for_P1)
paste0("Total entries: ", original)
spr_df_for_P1 <- spr_df_for_P1 %>%
    filter(Reading.time > 100)
afterFilter <- nrow(spr_df_for_P1)
paste0("Total entries with less than 100 ms response time for ROI: ", original - afterFilter)
```
```{r}
# Remove responses greater than 5000ms
original <- nrow(spr_df_for_P1)
spr_df_for_P1 <- spr_df_for_P1 %>%
    filter(Reading.time < 5000)
afterFilter <- nrow(spr_df_for_P1)
paste0("Total entries with greater than 5000 ms response time for ROI: ", original - afterFilter)
```
```{r}
# Participant level exclusion: Any participant whose mean RT for ROI is greater than 2.5SD of the sample pool will be discarded
# It is better to use the log RT instead of RT. Anyway, it is necessary for our modelling too!
spr_df_for_P1$p1LogRT <- log(spr_df_for_P1$Reading.time)
participant_means <- spr_df_for_P1 %>%
    group_by(user_id) %>%
    summarize(mean_RT = mean(p1LogRT, na.rm = TRUE))

meanRT <- mean(participant_means$mean_RT)
sdRT <- sd(participant_means$mean_RT)
threshold <- meanRT + 2.5 * sdRT
participants_to_remove <- participant_means %>%
    filter(mean_RT > threshold) %>%
    select(user_id)

spr_df_for_P1 <- spr_df_for_P1 %>%
    filter(!user_id %in% participants_to_remove$user_id)

paste0("Total participants discarded based on exclusion criteria: ", unique(nrow(participants_to_remove)))
```
```{r}
# Next criterion: Any responses to an item which is greater than 2.5SD from the mean will be discarded.
# This is to remove the outlier items that are extremely hard for everyone
item_means <- spr_df_for_P1 %>%
    group_by(ItemID) %>%
    summarize(mean_Item_RT = mean(p1LogRT, na.rm = TRUE))

meanItemRT <- mean(item_means$mean_Item_RT)
sdItemRT <- sd(item_means$mean_Item_RT)
threshold <- meanItemRT + 2.5 * sdItemRT
items_to_remove <- item_means %>%
    filter(mean_Item_RT > threshold) %>%
    select(ItemID)

spr_df_for_P1 <- spr_df_for_P1 %>%
    filter(!ItemID %in% items_to_remove$ItemID)
paste0("Total items discarded: based on exclusion criteria: ", unique(nrow(items_to_remove)))
```
```{r}
# Factoring and defining the contrasts
spr_df_for_P1$WordTypeP1 <- as.factor(spr_df_for_P1$WordTypeP1)
# This contrast assigns 1 to C1 and -1 to IH.
contrasts(spr_df_for_P1$WordTypeP1) <- contr.sum(2)
```
```{r}
# Center and scale
spr_df_for_P1$orthographic_distance_z <- scale(spr_df_for_P1$orthographic_distance, scale = TRUE, center = TRUE)
spr_df_for_P1$phonetic_distance_z <- scale(spr_df_for_P1$phonetic_distance, scale = TRUE, center = TRUE)
spr_df_for_P1$LogFreqDE_z <- scale(spr_df_for_P1$LogFreqDE, scale = TRUE, center = TRUE)
spr_df_for_P1$LogFreqUS_z <- scale(spr_df_for_P1$LogFreqUS, scale = TRUE, center = TRUE)
spr_df_for_P1$wordLengthP1 <- nchar(spr_df_for_P1$Value)
spr_df_for_P1$wordLengthP1_z <- scale(spr_df_for_P1$wordLengthP1, scale = TRUE, center = TRUE)
spr_df_for_P1$LexTALE_z <- scale(spr_df_for_P1$LexTALE, scale = TRUE, center = TRUE)
```
```{r}
# Quick check of distribution of reading times
hist(spr_df_for_P1$Reading.time)
hist(spr_df_for_P1$p1LogRT)

# QQ plot of log RT
qqnorm(spr_df_for_P1$p1LogRT)
qqline(spr_df_for_P1$p1LogRT, col = "red")
```
```{r}
# Save the dataframe for future references
write.csv(spr_df_for_P1, "D:/Studies/Thesis/Analysis/Inhibition-Control-Bilingual-Sentence-Processing/data/final analysis/P1_ROI.csv", row.names = FALSE)
```
```{r}
# ============================================================
# 1a. ROI Analysis - Confirmatory Model
# ============================================================
```
```{r}
base_formula_ROI <- as.formula(
    "p1LogRT ~ WordTypeP1 * inhibition_control_score + LogFreqDE_z * WordTypeP1 + LogFreqUS_z + wordLengthP1_z + (1 + WordTypeP1 | user_id) + (1 + WordTypeP1 | ItemID)"
)
# Model fit
base_model_ROI <- lmerTest::lmer(formula = base_formula_ROI, data = spr_df_for_P1)
summary(base_model_ROI)
```
```{r}
# Removing LogFreqDE_z:WordTypeP1
model_without_LogFreqDE_z_WordTypeP1 <- update(base_model_ROI, . ~ . - LogFreqDE_z:WordTypeP1)
summary(model_without_LogFreqDE_z_WordTypeP1)
anova(model_without_LogFreqDE_z_WordTypeP1, base_model_ROI)
```
```{r}
# Removing LogFreqDE_z
model_without_LogFreqDE_z <- update(model_without_LogFreqDE_z_WordTypeP1, . ~ . - LogFreqDE_z)
anova(model_without_LogFreqDE_z_WordTypeP1, model_without_LogFreqDE_z)
# Removing LogFreqDE_z was useful. Explore with this model.
```
```{r}
model_without_LogFreqUS_z <- update(model_without_LogFreqDE_z, . ~ . - LogFreqUS_z)
anova(model_without_LogFreqDE_z, model_without_LogFreqUS_z)
# Removing LogFreqUS_z was useful. Explore with this model.
```
```{r}
model_without_wordLength_z <- update(model_without_LogFreqUS_z, . ~ . - wordLengthP1_z)
anova(model_without_wordLength_z, model_without_LogFreqUS_z)
# Removing wordLengthP1_z was not effective. So switch back to model_without_LogFreqUS_z
```
```{r}
# Remove The random slope of P1 for items
final_model_roi_base <- update(model_without_LogFreqUS_z, . ~ . - (1 + WordTypeP1 | ItemID) + (1 | ItemID))
anova(final_model_roi_base, model_without_LogFreqUS_z)
summary(final_model_roi_base)
```
```{r}
# Remove The random slope of P1 for user
final_model_roi_base_without_slopes <- update(final_model_roi_base, . ~ . - (1 + WordTypeP1 | user_id) + (1 | user_id))
anova(final_model_roi_base_without_slopes, final_model_roi_base)
summary(final_model_roi_base_without_slopes)
```
```{r}
# Everything is done! Check the summary
summary(model_without_LogFreqUS_z)
AIC(final_model_roi_base_without_slopes)
BIC(final_model_roi_base_without_slopes)
saveRDS(final_model_roi_base_without_slopes, "P1_ROI_Base_lmer_no_slopes.rds")
```
```{r}
# ============================================================
# 1a. ROI Analysis - Exploratory Model
# ============================================================
```
```{r}
extended_base_formula_ROI <- as.formula(
    "p1LogRT ~  WordTypeP1 * inhibition_control_score * LexTALE_z + LogFreqDE_z*WordTypeP1 + LogFreqUS_z + phonetic_distance_z * WordTypeP1 + orthographic_distance_z * WordTypeP1 + LexTALE_z * LogFreqUS_z + wordLengthP1_z + (1 + WordTypeP1  | user_id) + (1 + WordTypeP1 | ItemID)"
)
# Model fit
extended_base_model_ROI <- lmerTest::lmer(formula = extended_base_formula_ROI, data = spr_df_for_P1)
summary(extended_base_model_ROI)
# orthographic_distance_z is not significant,so remove it and start the ANOVA comparison
```
```{r}
model_without_orthographic_distance_z_P1 <- update(extended_base_model_ROI, . ~ . - orthographic_distance_z:WordTypeP1)
anova(extended_base_model_ROI, model_without_orthographic_distance_z_P1)
# Removing improved the model fit
```
```{r}
model_without_orthographic_distance_z <- update(model_without_orthographic_distance_z_P1, . ~ . - orthographic_distance_z)
anova(model_without_orthographic_distance_z_P1, model_without_orthographic_distance_z)
# Remove the orthographic distance from original model was useful. Now remove the phonetic distance
```
```{r}
model_without_phonetic_distance_z_P1 <- update(model_without_orthographic_distance_z, . ~ . - phonetic_distance_z:WordTypeP1)
anova(model_without_phonetic_distance_z_P1, model_without_orthographic_distance_z_P1)
# Now remove the phonetic_distance_z.
```
```{r}
model_without_phonetic_distance_z <- update(model_without_phonetic_distance_z_P1, . ~ . - phonetic_distance_z)
anova(model_without_phonetic_distance_z, model_without_phonetic_distance_z_P1)
# Removing the phonetic_distance_z was useful.
```
```{r}
# Look at the summary to see if there are other insignificant fixed effects that are worth removing
summary(model_without_phonetic_distance_z)
# Remove WordTypeP11:LogFreqDE_z

model_without_LogFreqDE_z_P1 <- update(model_without_phonetic_distance_z, . ~ . - WordTypeP1:LogFreqDE_z)
anova(model_without_LogFreqDE_z_P1, model_without_phonetic_distance_z)
# Removing WordTypeP11:LogFreqDE_z was useful. Now remove LogFreqDE_z
```
```{r}
model_without_LogFreqDE_z <- update(model_without_LogFreqDE_z_P1, . ~ . - LogFreqDE_z)
anova(model_without_LogFreqDE_z_P1, model_without_LogFreqDE_z)
```
```{r}
model_without_LogFreqUS_z_LexTALE <- update(model_without_LogFreqDE_z, . ~ . - LexTALE_z:LogFreqUS_z)
anova(model_without_LogFreqUS_z_LexTALE, model_without_LogFreqDE_z)
# model_without_LogFreqUS_z_LexTALE is better
# Check the model summary
summary(model_without_LogFreqUS_z_LexTALE)
```
```{r}
model_without_LogFreqUS_z <- update(model_without_LogFreqUS_z_LexTALE, . ~ . - LogFreqUS_z)
anova(model_without_LogFreqUS_z, model_without_LogFreqUS_z_LexTALE)
# model_without_LogFreqUS_z is better
```
```{r}
# Remove The random slope of P1 for items
final_model_roi_base <- update(model_without_LogFreqUS_z, . ~ . - (1 + WordTypeP1 | ItemID) + (1 | ItemID))
anova(final_model_roi_base, model_without_LogFreqUS_z)
summary(final_model_roi_base)
```
```{r}
# Remove The random slope of P1 for items
final_model_roi_exp_without_slopes <- update(final_model_roi_base, . ~ . - (1 + WordTypeP1 | user_id) + (1 | user_id))
anova(final_model_roi_exp_without_slopes, final_model_roi_base)
summary(final_model_rofinal_model_roi_exp_without_slopesi_exp_without_slopes)
```

```{r}
summary(final_model_roi_exp_without_slopes)
saveRDS(final_model_roi_exp_without_slopes, "P1_ROI_Exp_lmer_no_slopes.rds")
# Nothing more can be removed from this model.
# Just a marginal negative two-way interaction in case of inhibition_control_score:LexTALE_z
```
```{r}
# ============================================================
# 2. Spillover 1 Analysis
# ============================================================
```
```{r}
spr_df_for_SO1 <- spr_df_filtered %>%
    left_join(materials_df %>% select(SentenceID, Critical.Position.Index, WordatP1),
        by = "SentenceID"
    ) %>%
    filter(Parameter == Critical.Position.Index + 1)
```
nrow(spr_df_filtered)
```{r}
# Remove responses lesser than 100ms
original <- nrow(spr_df_for_SO1)
paste0("Total entries: ", original)
spr_df_for_SO1 <- spr_df_for_SO1 %>%
    filter(Reading.time > 100)

afterFilter <- nrow(spr_df_for_SO1)
paste0("Total entries with less than 100 ms response time for SO1: ", original - afterFilter)
```
```{r}
# Remove responses greater than 5000ms
original <- nrow(spr_df_for_SO1)
spr_df_for_SO1 <- spr_df_for_SO1 %>%
    filter(Reading.time < 5000)

afterFilter <- nrow(spr_df_for_SO1)
paste0("Total entries with greater than 5000 ms response time for SO1: ", original - afterFilter)
```
```{r}
# Participant level exclusion: Any participant whose mean RT for ROI is greater than 2.5SD of the sample pool will be discarded
# It is better to use the log RT instead of RT. Anyway, it is necessary for our modelling too!
spr_df_for_SO1$p1LogRT <- log(spr_df_for_SO1$Reading.time)
participant_means <- spr_df_for_SO1 %>%
    group_by(user_id) %>%
    summarize(mean_RT = mean(p1LogRT, na.rm = TRUE))

meanRT <- mean(participant_means$mean_RT)
sdRT <- sd(participant_means$mean_RT)
threshold <- meanRT + 2.5 * sdRT
participants_to_remove <- participant_means %>%
    filter(mean_RT > threshold) %>%
    select(user_id)

spr_df_for_SO1 <- spr_df_for_SO1 %>%
    filter(!user_id %in% participants_to_remove$user_id)

paste0("Total participants discarded after exclusion criteria: ", unique(nrow(participants_to_remove)))
```
```{r}
# Next criterion: Any responses to an item which is greater than 2.5SD from the mean will be discarded.
# This is to remove the outlier items that are extremely hard for everyone
item_means <- spr_df_for_SO1 %>%
    group_by(ItemID) %>%
    summarize(mean_Item_RT = mean(p1LogRT, na.rm = TRUE))

meanItemRT <- mean(item_means$mean_Item_RT)
sdItemRT <- sd(item_means$mean_Item_RT)
threshold <- meanItemRT + 2.5 * sdItemRT
items_to_remove <- item_means %>%
    filter(mean_Item_RT > threshold) %>%
    select(ItemID)

spr_df_for_SO1 <- spr_df_for_SO1 %>%
    filter(!ItemID %in% items_to_remove$ItemID)
paste0("Total items discarded after exclusion criteria: ", unique(nrow(items_to_remove)))
```
```{r}
# Factoring and defining the contrasts
spr_df_for_SO1$WordTypeP1 <- as.factor(spr_df_for_SO1$WordTypeP1)
contrasts(spr_df_for_SO1$WordTypeP1) <- contr.sum(2)
```
```{r}
# Add the Spillover word English frequency for addressing variance
matchesValue <- match(toupper(spr_df_for_SO1$Value), toupper(subtlexus$Word))
spr_df_for_SO1$FreqSO1 <- subtlexus$FREQlow[matchesValue]
spr_df_for_SO1$LogFreqSO1 <- log10(spr_df_for_SO1$FreqSO1) + 1
```
```{r}
spr_df_for_SO1$orthographic_distance_z <- scale(spr_df_for_SO1$orthographic_distance, scale = TRUE, center = TRUE)
spr_df_for_SO1$phonetic_distance_z <- scale(spr_df_for_SO1$phonetic_distance, scale = TRUE, center = TRUE)
spr_df_for_SO1$LogFreqDE_z <- scale(spr_df_for_SO1$LogFreqDE, scale = TRUE, center = TRUE)
spr_df_for_SO1$LogFreqUS_z <- scale(spr_df_for_SO1$LogFreqUS, scale = TRUE, center = TRUE)
spr_df_for_SO1$wordLengthP1 <- nchar(spr_df_for_SO1$Value)
spr_df_for_SO1$wordLengthP1_z <- scale(spr_df_for_SO1$wordLengthP1, scale = TRUE, center = TRUE)
spr_df_for_SO1$LexTALE_z <- scale(spr_df_for_SO1$LexTALE, scale = TRUE, center = TRUE)
spr_df_for_SO1$LogFreqSO1_z <- scale(spr_df_for_SO1$LogFreqSO1, scale = TRUE, center = TRUE)
```
```{r}
# Save the dataframe for future references
write.csv(spr_df_for_SO1, "D:/Studies/Thesis/Analysis/Inhibition-Control-Bilingual-Sentence-Processing/data/final analysis/P1_SO1.csv", row.names = FALSE)
```
```{r}
# ============================================================
# 2a. Spillover 1 Analysis - Confirmatory Model
# ============================================================
```
```{r}
base_formula_ROI <- as.formula(
    "p1LogRT ~ WordTypeP1 * inhibition_control_score + LogFreqDE_z * WordTypeP1 + LogFreqUS_z + LogFreqSO1_z + wordLengthP1_z + (1 + WordTypeP1  | user_id) + (1 + WordTypeP1 | ItemID)"
)
base_model_ROI_SO1 <- lmerTest::lmer(formula = base_formula_ROI, data = spr_df_for_SO1)
summary(base_model_ROI_SO1)
```
```{r}
# Remove WordTypeP11:LogFreqDE_z
model_without_WordTypeP1_LogFreqDE_z <- update(base_model_ROI_SO1, . ~ . - WordTypeP1:LogFreqDE_z)
anova(base_model_ROI_SO1, model_without_WordTypeP1_LogFreqDE_z)
```
```{r}
# Base model looks solid already. I can remove LogFreqSO1_z
model_without_LogFreqSO1_z <- update(model_without_WordTypeP1_LogFreqDE_z, . ~ . - LogFreqSO1_z)
anova(model_without_LogFreqSO1_z, model_without_WordTypeP1_LogFreqDE_z)
```
```{r}
# Remove the slopes
final_model_without_P1_rs <- update(model_without_LogFreqSO1_z, . ~ . - (1 + WordTypeP1 | ItemID) + (1 | ItemID))
anova(final_model_without_P1_rs, model_without_LogFreqSO1_z)
```
```{r}
# Remove the slopes
final_model_without_P1_rs_us <- update(final_model_without_P1_rs, . ~ . - (1 + WordTypeP1 | user_id) + (1 | user_id))
anova(final_model_without_P1_rs_us, final_model_without_P1_rs)
```
```{r}
# Stopping here!
saveRDS(final_model_without_P1_rs_us, "P1_SO1_Base_lmer_no_slopes.rds")
summary(final_model_without_P1_rs_us)
```
```{r}
# ============================================================
# 2b. Spillover 1 Analysis - Exploratory Model
# ============================================================
```
```{r}
# Now plug in extra variables as part of exploratory analyses
extended_base_formula_ROI_SO1 <- as.formula(
    "p1LogRT ~ WordTypeP1 * inhibition_control_score + WordTypeP1 * inhibition_control_score * LexTALE_z + LogFreqSO1_z + LexTALE_z * inhibition_control_score + LogFreqDE_z * WordTypeP1 + LogFreqUS_z + phonetic_distance_z * WordTypeP1 + orthographic_distance_z * WordTypeP1 + LexTALE_z * LogFreqUS_z + wordLengthP1_z + (1 + WordTypeP1  | user_id) + (1 + WordTypeP1 | ItemID)"
)
# Model fit
extended_base_model_ROI_SO1 <- lmerTest::lmer(formula = extended_base_formula_ROI_SO1, data = spr_df_for_SO1)
summary(extended_base_model_ROI_SO1)
# Phonetic distance, Orthographic distance are not significant. Good candidates to remove
# Start with P11:orthographic_distance_z
```
```{r}
model_without_P11_orthographic_distance_z <- update(extended_base_model_ROI_SO1, . ~ . - WordTypeP1:orthographic_distance_z)
anova(model_without_P11_orthographic_distance_z, extended_base_model_ROI_SO1)
# model_without_P11_orthographic_distance_z is better. Now remove the distance by itself
```
```{r}
model_without_orthographic_distance_z <- update(model_without_P11_orthographic_distance_z, . ~ . - orthographic_distance_z)
anova(model_without_P11_orthographic_distance_z, model_without_orthographic_distance_z)
# The new model is better. Now remove the phonetic distance
```
```{r}
model_without_phonetic_distance_z_P1 <- update(model_without_orthographic_distance_z, . ~ . - WordTypeP1:phonetic_distance_z)
anova(model_without_orthographic_distance_z, model_without_phonetic_distance_z_P1)
# Phonetic distance: P1 was important, but the distance alone may not be so. Just remove the main effect now from model_without_orthographic_distance_z
```
```{r}
model_without_phonetic_distance_z <- update(model_without_phonetic_distance_z_P1, . ~ . - phonetic_distance_z)
anova(model_without_phonetic_distance_z_P1, model_without_phonetic_distance_z)
```
```{r}
# Remove the interaction of WordTypeP1 and LogFreqDE_z
model_without_P1_LogFreqDE_z <- update(model_without_phonetic_distance_z, . ~ . - WordTypeP1:LogFreqDE_z)
anova(model_without_phonetic_distance_z, model_without_P1_LogFreqDE_z)
```
```{r}
# Remove LogFreqSO1_z
model_without_LogFreqSO1_z <- update(model_without_P1_LogFreqDE_z, . ~ . - LogFreqSO1_z)
anova(model_without_orthographic_distance_z, model_without_LogFreqSO1_z)
```
```{r}
# Remove item random slope
fin_model_without_item_rs <- update(model_without_LogFreqSO1_z, . ~ . - (1 + WordTypeP1 | ItemID) + (1 | ItemID))
anova(model_without_LogFreqSO1_z, fin_model_without_item_rs)
```
```{r}
# Remove user random slope
fin_model_without_user_rs <- update(fin_model_without_item_rs, . ~ . - (1 + WordTypeP1 | user_id) + (1 | user_id))
anova(fin_model_without_user_rs, fin_model_without_item_rs)
```
```{r}
# Nothing more to improve
summary(fin_model_without_user_rs)
saveRDS(fin_model_without_user_rs, "P1_SO1_Exp_lmer_no_slopes.rds")
```
```{r}
# ============================================================
# 2. Spillover 2 Analysis
# ============================================================
```
```{r}
spr_df_for_SO2 <- spr_df_filtered %>%
    left_join(materials_df %>% select(SentenceID, Critical.Position.Index, WordatP1),
        by = "SentenceID"
    ) %>%
    filter(Parameter == Critical.Position.Index + 2)
```
```{r}
# Remove responses lesser than 100ms
original <- nrow(spr_df_for_SO2)
paste0("Total entries", original)
spr_df_for_SO2 <- spr_df_for_SO2 %>%
    filter(Reading.time > 100)

afterFilter <- nrow(spr_df_for_SO2)
paste0("Total entries with less than 100 ms response time for SO2: ", original - afterFilter)
```
```{r}
# Remove responses greater than 5000ms
original <- nrow(spr_df_for_SO2)
spr_df_for_SO2 <- spr_df_for_SO2 %>%
    filter(Reading.time < 5000)
afterFilter <- nrow(spr_df_for_SO2)
paste0("Total entries with greater than 5000 ms response time for SO2: ", original - afterFilter)
```
```{r}
# Participant level exclusion: Any participant whose mean RT for ROI is greater than 2.5SD of the sample pool will be discarded
spr_df_for_SO2$p1LogRT <- log(spr_df_for_SO2$Reading.time)

participant_means <- spr_df_for_SO2 %>%
    group_by(user_id) %>%
    summarize(mean_RT = mean(p1LogRT, na.rm = TRUE))

meanRT <- mean(participant_means$mean_RT)
sdRT <- sd(participant_means$mean_RT)
threshold <- meanRT + 2.5 * sdRT
participants_to_remove <- participant_means %>%
    filter(mean_RT > threshold) %>%
    select(user_id)

spr_df_for_SO2 <- spr_df_for_SO2 %>%
    filter(!user_id %in% participants_to_remove$user_id)

paste0("Total participants discarded aftere exclusion criteria: ", unique(nrow(participants_to_remove)))
```
```{r}
# Next criterion: Any responses to an item which is greater than 2.5SD from the mean will be discarded.
# This is to remove the outlier items that are extremely hard for everyone
item_means <- spr_df_for_SO2 %>%
    group_by(ItemID) %>%
    summarize(mean_Item_RT = mean(p1LogRT, na.rm = TRUE))

meanItemRT <- mean(item_means$mean_Item_RT)
sdItemRT <- sd(item_means$mean_Item_RT)
threshold <- meanItemRT + 2.5 * sdItemRT
items_to_remove <- item_means %>%
    filter(mean_Item_RT > threshold) %>%
    select(ItemID)

spr_df_for_SO2 <- spr_df_for_SO2 %>%
    filter(!ItemID %in% items_to_remove$ItemID)
paste0("Total items discarded after exclusion criteria: ", unique(nrow(items_to_remove)))
```
```{r}
# Factoring and defining the contrasts
spr_df_for_SO2$WordTypeP1 <- as.factor(spr_df_for_SO2$WordTypeP1)
contrasts(spr_df_for_SO2$WordTypeP1) <- contr.sum(2)
```
```{r}
# Add the Spillover word English frequency for addressing variance
matchesValue <- match(toupper(spr_df_for_SO2$Value), toupper(subtlexus$Word))
spr_df_for_SO2$FreqSO2 <- subtlexus$FREQlow[matchesValue]
spr_df_for_SO2$LogFreqSO2 <- log10(spr_df_for_SO2$FreqSO2) + 1
```
```{r}
spr_df_for_SO2$orthographic_distance_z <- scale(spr_df_for_SO2$orthographic_distance, scale = TRUE, center = TRUE)
spr_df_for_SO2$phonetic_distance_z <- scale(spr_df_for_SO2$phonetic_distance, scale = TRUE, center = TRUE)
spr_df_for_SO2$LogFreqDE_z <- scale(spr_df_for_SO2$LogFreqDE, scale = TRUE, center = TRUE)
spr_df_for_SO2$LogFreqUS_z <- scale(spr_df_for_SO2$LogFreqUS, scale = TRUE, center = TRUE)
spr_df_for_SO2$wordLengthP1 <- nchar(spr_df_for_SO2$Value)
spr_df_for_SO2$wordLengthP1_z <- scale(spr_df_for_SO2$wordLengthP1, scale = TRUE, center = TRUE)
spr_df_for_SO2$LexTALE_z <- scale(spr_df_for_SO2$LexTALE, scale = TRUE, center = TRUE)
spr_df_for_SO2$LogFreqSO2_z <- scale(spr_df_for_SO2$LogFreqSO2, scale = TRUE, center = TRUE)
```
```{r}
# ============================================================
# 3a. Spillover 2 Analysis - Confirmatory Model
# ============================================================
```
```{r}
# Base formula is only concerned with main hypotheses
base_formula_ROI_SO2 <- as.formula(
    "p1LogRT ~ WordTypeP1 * inhibition_control_score + LogFreqDE_z * WordTypeP1 + LogFreqUS_z + LogFreqSO2_z + wordLengthP1_z + (1 + WordTypeP1  | user_id) + (1 + WordTypeP1 | ItemID)"
)
# Model fit
base_model_ROI_SO2 <- lmerTest::lmer(formula = base_formula_ROI_SO2, data = spr_df_for_SO2)
summary(base_model_ROI_SO2)
```
```{r}
# Now remove LogFreqUS_z from base_model_ROI_SO2
model_without_LogFreqUS_z <- update(base_model_ROI_SO2, . ~ . - LogFreqUS_z)
anova(model_without_LogFreqUS_z, base_model_ROI_SO2)
```
```{r}
# Remove WordTypeP1:LogFreqDE_z
model_without_WordTypeP1_LogFreqDE_z <- update(model_without_LogFreqUS_z, . ~ . - WordTypeP1:LogFreqDE_z)
anova(model_without_LogFreqUS_z, model_without_WordTypeP1_LogFreqDE_z)
```
```{r}
model_without_LogFreqDE_z <- update(model_without_WordTypeP1_LogFreqDE_z, . ~ . - LogFreqDE_z)
anova(model_without_LogFreqDE_z, model_without_WordTypeP1_LogFreqDE_z)
# model_without_WordTypeP1_LogFreqDE_z is marginally better. Use it!
```
```{r}
# remove word length
model_without_wordlength <- update(model_without_WordTypeP1_LogFreqDE_z, . ~ . - wordLengthP1_z)
anova(model_without_WordTypeP1_LogFreqDE_z, model_without_wordlength)
```
```{r}
# Remove item random slope
fin_model_without_item_rs <- update(model_without_wordlength, . ~ . - (1 + WordTypeP1 | ItemID) + (1 | ItemID))
anova(model_without_wordlength, fin_model_without_item_rs)
```
```{r}
# Remove user random slope
fin_model_without_user_rs <- update(fin_model_without_item_rs, . ~ . - (1 + WordTypeP1 | user_id) + (1 | user_id))
anova(fin_model_without_user_rs, fin_model_without_item_rs) # fin_model_without_user_rs is the best model
```
```{r}
# Remove LogFreqDE_z again
model_without_LogFreqDE_z <- update(fin_model_without_user_rs, . ~ . - LogFreqDE_z)
anova(model_without_LogFreqDE_z, fin_model_without_user_rs)
```
```{r}
# fin_model_without_user_rs is still the better model!
summary(fin_model_without_user_rs)
saveRDS(fin_model_without_user_rs, "P1_SO2_Base_lmer_no_slopes.rds")
```
```{r}
# ============================================================
# 3b. Spillover 2 Analysis - Exploratory Model
# ============================================================
```
```{r}
extended_base_formula_ROI_SO2 <- as.formula(
    "p1LogRT ~ WordTypeP1 * inhibition_control_score * LexTALE_z + LexTALE_z * inhibition_control_score + LogFreqDE_z * WordTypeP1 + LogFreqUS_z + LogFreqSO2_z + phonetic_distance_z * WordTypeP1 + orthographic_distance_z * WordTypeP1 + LexTALE_z * LogFreqUS_z + wordLengthP1_z + (1 + WordTypeP1  | user_id) + (1 + WordTypeP1 | ItemID)"
)
# Model fit
extended_base_model_ROI_SO2 <- lmerTest::lmer(formula = extended_base_formula_ROI_SO2, data = spr_df_for_SO2)
summary(extended_base_model_ROI_SO2)
```
```{r}
# Remove phonetic_distance_z : WordTypeP1
model_without_phonetic_distance_z_P1 <- update(extended_base_model_ROI_SO2, . ~ . - phonetic_distance_z:WordTypeP1)
anova(extended_base_model_ROI_SO2, model_without_phonetic_distance_z_P1)
```
```{r}
# Now remove the main effect of phonetic_distance_z
model_without_phonetic_distance_z <- update(model_without_phonetic_distance_z_P1, . ~ . - phonetic_distance_z)
anova(model_without_phonetic_distance_z, model_without_phonetic_distance_z_P1)
```
```{r}
# Removing wordLength
model_without_wordLength <- update(model_without_phonetic_distance_z, . ~ . - wordLengthP1_z)
summary(model_without_wordLength)
anova(model_without_wordLength, model_without_phonetic_distance_z)
```
```{r}
# Remove orthographic_distance_z : WordTypeP1
model_without_orthographic_distance_z_P1 <- update(model_without_wordLength, . ~ . - orthographic_distance_z:WordTypeP1)
summary(model_without_orthographic_distance_z_P1)
anova(model_without_wordLength, model_without_orthographic_distance_z_P1)
```
```{r}
# Now remove the main effect of orthographic_distance_z
model_without_orthographic_distance_z <- update(model_without_orthographic_distance_z_P1, . ~ . - orthographic_distance_z)
anova(model_without_orthographic_distance_z, model_without_orthographic_distance_z_P1)
summary(model_without_orthographic_distance_z)
```
```{r}
# Remove LogFreqUS_z:LexTALE_z
model_without_LogFreqUS_z_LexTALE_z <- update(model_without_orthographic_distance_z, . ~ . - LogFreqUS_z:LexTALE_z)
summary(model_without_LogFreqUS_z_LexTALE_z)
anova(model_without_LogFreqUS_z_LexTALE_z, model_without_orthographic_distance_z)
```
```{r}
# Now remove LogFreqUS_z
model_without_LogFreqUS_z <- update(model_without_LogFreqUS_z_LexTALE_z, . ~ . - LogFreqUS_z)
summary(model_without_LogFreqUS_z)
anova(model_without_LogFreqUS_z_LexTALE_z, model_without_LogFreqUS_z)
```
```{r}
# Now remove the LogFeqDE_z:WordTypeP1
model_without_LogFreqDE_z_P1 <- update(model_without_LogFreqUS_z, . ~ . - WordTypeP1:LogFreqDE_z)
anova(model_without_LogFreqDE_z_P1, model_without_LogFreqUS_z)
```
```{r}
model_without_LogFreqDE_z <- update(model_without_LogFreqDE_z_P1, . ~ . - LogFreqDE_z)
anova(model_without_LogFreqDE_z_P1, model_without_LogFreqDE_z)
# model_without_LogFreqDE_z_P1 is better
```
```{r}
# Slope Removals
fin_model_without_item_rs <- update(model_without_LogFreqDE_z_P1, . ~ . - (1 + WordTypeP1 | ItemID) + (1 | ItemID))
anova(model_without_LogFreqDE_z_P1, fin_model_without_item_rs)
```
```{r}
fin_model_without_user_rs <- update(fin_model_without_item_rs, . ~ . - (1 + WordTypeP1 | user_id) + (1 | user_id))
anova(fin_model_without_user_rs, fin_model_without_item_rs)
```
```{r}
# fin_model_without_user_rs is the best model
summary(fin_model_without_user_rs)
saveRDS(fin_model_without_user_rs, "P1_SO2_Exp_lmer_no_slopes.rds")
```